canaccessR: An open data product for analyzing transportation accessibility to employment and grocery stores in Canada's largest metropolitan areas.

### 2025-03-07 paper/Manuscript-Data-Package.Rmd
minor updates
minor updates to draft; descriptive statistics and how to use still undone.

The locations included in the data package comprise the 12 largest (population-wise) Census metropolitan areas (CMA's) based on the 2021 Canadian Census [@governmentofcanada2021CensusPopulation2021] [^1], which are Toronto, Montreal, Vancouver, Ottawa-Gatineau, Calgary, Edmonton, Quebec City, Winnipeg, Hamilton, Kitchener-Cambridge-Waterloo, London, and Halifax.

[^1]: We included Oshawa, Ontario, as part of the Greater Toronto Area (GTA) because of the former's proximity to the latter. We also included Abbotsford-Mission, British Columbia, as part of the Vancouver metropolitan area because of the former's proximity to a transit station on the region’s West Coast Express commuter rail line.

We used four main data sources to construct the {canaccesR} data package: General Transit Feed Specification (GTFS), OpenStreetMap (OSM), DMTI’s Enhanced Points of Interest, and Statistics Canada Census data.

<!-- ## General Transit Feed Specification -->
The GTFS files from the transit agencies from all the related areas provided the information on public transit schedule and their changes from 2019 to 2023.

<!-- ## OpenStreetMaps -->
The OpenStreetMap data provided information on the transit network and was necessary for the routing used to estimate the travel time matrices.

<!-- ## Census Data -->
The 2016 and 2021 Statistics' Canada Census data provided the information on the spatial distribution of the population and the number of workplace locations, which represent the quantity of employment across the CMA's [@governmentofcanada2016CensusPopulation2016, governmentofcanada2021CensusPopulation2021].
All of the resulting travel time matrices were aggregated at the Dissemination Area (DA) level, which comprise the fundamental unit of analysis in data package.

<!-- ## DMTI’s Enhanced Points of Interest -->
The DMTI’s Enhanced Points of Interest

@@

Below, we present some of the basic statistics of

@@

In this paper, we describe the {canaccesR} data package, created using the {r5r} R package and transit schedule, street network, employment, and population data.
The package's main contents refers to the ready-to-use travel time matrices for public transit to reach employment and groceries stores in Canada's 12 largest urban areas.
We expect the contents of the package to be used in transportation accessibility evaluations within and across those regions.
Moreover, these datasets can be used in further equity assessments that evaluate the distribution of accessibility across space and between social groups.
In other words, we hope that by making these datasets publicly available, future analysis can contribute to making Canada's transportation system more just and fair, considering accessibility's as the main social good of transportation [MARTENS JUSTICE], and the inherent connection between public transit and the "right to the city" [COGGIN].

>>>>

The locations included in the data package comprise the 12 largest (population-wise) Census metropolitan areas (CMA's) based on the 2021 Canadian Census [@governmentofcanada2021CensusPopulation2021] [^1]. 
These locations are Toronto, Montreal, Vancouver, Ottawa-Gatineau, Calgary, Edmonton, Quebec City, Winnipeg, Hamilton, Kitchener-Cambridge-Waterloo, London, and Halifax.

[^1]: We included Oshawa, Ontario, as part of the Greater Toronto Area (GTA) because of the former's proximity to the latter. We also included Abbotsford-Mission, British Columbia, as part of the Vancouver metropolitan area because of the former's proximity to a transit station on the region’s West Coast Express commuter rail line.
We used four main data sources to construct the {canaccesR} data package: General Transit Feed Specification (GTFS), OpenStreetMap (OSM), DMTI’s Enhanced Points of Interest, and Statistics Canada Census data.

<!-- ## General Transit Feed Specification -->
We used the GTFS files from all transit agencies within the selected CMA's to obtain the information on their public transit schedule from 2019 to 2023.
<!-- ## OpenStreetMaps -->
The OpenStreetMap data provided information on the transit network and was necessary for the routing used to estimate the travel time matrices.
<!-- ## Census Data -->
We used the 2016 and 2021 Canadian Census' information on the spatial distribution of the population and the number of workplace locations (employment) across the CMA's [@governmentofcanada2016CensusPopulation2016, governmentofcanada2021CensusPopulation2021].
We aggregayed all the resulting travel time matrices at the Dissemination Area (DA) level, which comprise the fundamental unit of analysis in data package.
<!-- ## DMTI’s Enhanced Points of Interest -->
Finally, we use the DMTI’s Enhanced Points of Interest dataset to obtain the location of the groceries stores within every urban area selected [@dmtispatialincEnhancedPointsInterest2015]. 
We filtered the locations within the DMTI dataset using the grocery stores code from the North American Industry Classification System (NAICS) and the Standard Industrial Classification (SIC).

@@

Below, we present some of the basic statistics from the travel time matrices contained in the {canaccessR} package.

@@

In this paper, we describe the {canaccesR} data package, created using the {r5r} R package and transit schedule, street network, employment, and population data.
The package's main contents refers to the ready-to-use travel time matrices for public transit to reach employment and groceries stores in Canada's 12 largest urban areas.
We expect the contents of the package to be used in transportation accessibility evaluations within and across those regions.
Moreover, these datasets can be used in further equity assessments that evaluate the distribution of accessibility across space and between social groups.
In other words, we hope that by making these datasets publicly available, future analysis can contribute to making Canada's transportation system more just and fair, considering accessibility's as the main social good of transportation [@martensTransportJusticeDesigning2016], and the inherent connection between public transit and the "right to the city" [@cogginRightTransportMoving2015].

### 2025-03-07 paper/Manuscript-Data-Package.tex
minor updates
minor updates to draft; descriptive statistics and how to use still undone.

The locations included in the data package comprise the 12 largest
(population-wise) Census metropolitan areas (CMA's) based on the 2021
Canadian Census \citep{governmentofcanada2021CensusPopulation2021}
\footnote{We included Oshawa, Ontario, as part of the Greater Toronto
  Area (GTA) because of the former's proximity to the latter. We also
  included Abbotsford-Mission, British Columbia, as part of the
  Vancouver metropolitan area because of the former's proximity to a
  transit station on the region's West Coast Express commuter rail line.},
which are Toronto, Montreal, Vancouver, Ottawa-Gatineau, Calgary,
Edmonton, Quebec City, Winnipeg, Hamilton, Kitchener-Cambridge-Waterloo,
London, and Halifax.

We used four main data sources to construct the \{canaccesR\} data
package: General Transit Feed Specification (GTFS), OpenStreetMap (OSM),
DMTI's Enhanced Points of Interest, and Statistics Canada Census data.

The GTFS files from the transit agencies from all the related areas
provided the information on public transit schedule and their changes
from 2019 to 2023.

The OpenStreetMap data provided information on the transit network and
was necessary for the routing used to estimate the travel time matrices.

The 2016 and 2021 Statistics' Canada Census data provided the
information on the spatial distribution of the population and the number
of workplace locations, which represent the quantity of employment
across the CMA's
\citep[governmentofcanada2021CensusPopulation2021]{governmentofcanada2016CensusPopulation2016}.
All of the resulting travel time matrices were aggregated at the
Dissemination Area (DA) level, which comprise the fundamental unit of
analysis in data package.

The DMTI's Enhanced Points of Interest

@@

Below, we present some of the basic statistics of

@@

In this paper, we describe the \{canaccesR\} data package, created using
the \{r5r\} R package and transit schedule, street network, employment,
and population data. The package's main contents refers to the
ready-to-use travel time matrices for public transit to reach employment
and groceries stores in Canada's 12 largest urban areas. We expect the
contents of the package to be used in transportation accessibility
evaluations within and across those regions. Moreover, these datasets
can be used in further equity assessments that evaluate the distribution
of accessibility across space and between social groups. In other words,
we hope that by making these datasets publicly available, future
analysis can contribute to making Canada's transportation system more
just and fair, considering accessibility's as the main social good of
transportation {[}MARTENS JUSTICE{]}, and the inherent connection
between public transit and the ``right to the city'' {[}COGGIN{]}.

>>>>

The locations included in the data package comprise the 12 largest
(population-wise) Census metropolitan areas (CMA's) based on the 2021
Canadian Census \citep{governmentofcanada2021CensusPopulation2021}
\footnote{We included Oshawa, Ontario, as part of the Greater Toronto
  Area (GTA) because of the former's proximity to the latter. We also
  included Abbotsford-Mission, British Columbia, as part of the
  Vancouver metropolitan area because of the former's proximity to a
  transit station on the region's West Coast Express commuter rail line.
  We used four main data sources to construct the \{canaccesR\} data
  package: General Transit Feed Specification (GTFS), OpenStreetMap
  (OSM), DMTI's Enhanced Points of Interest, and Statistics Canada
  Census data.}. These locations are Toronto, Montreal, Vancouver,
Ottawa-Gatineau, Calgary, Edmonton, Quebec City, Winnipeg, Hamilton,
Kitchener-Cambridge-Waterloo, London, and Halifax.

We used the GTFS files from all transit agencies within the selected
CMA's to obtain the information on their public transit schedule from
2019 to 2023. The OpenStreetMap data provided information on the transit
network and was necessary for the routing used to estimate the travel
time matrices. We used the 2016 and 2021 Canadian Census' information on
the spatial distribution of the population and the number of workplace
locations (employment) across the CMA's
\citep[governmentofcanada2021CensusPopulation2021]{governmentofcanada2016CensusPopulation2016}.
We aggregayed all the resulting travel time matrices at the
Dissemination Area (DA) level, which comprise the fundamental unit of
analysis in data package. Finally, we use the DMTI's Enhanced Points of
Interest dataset to obtain the location of the groceries stores within
every urban area selected
\citep{dmtispatialincEnhancedPointsInterest2015}. We filtered the
locations within the DMTI dataset using the grocery stores code from the
North American Industry Classification System (NAICS) and the Standard
Industrial Classification (SIC).

@@

Below, we present some of the basic statistics from the travel time
matrices contained in the \{canaccessR\} package.

@@

In this paper, we describe the \{canaccesR\} data package, created using
the \{r5r\} R package and transit schedule, street network, employment,
and population data. The package's main contents refers to the
ready-to-use travel time matrices for public transit to reach employment
and groceries stores in Canada's 12 largest urban areas. We expect the
contents of the package to be used in transportation accessibility
evaluations within and across those regions. Moreover, these datasets
can be used in further equity assessments that evaluate the distribution
of accessibility across space and between social groups. In other words,
we hope that by making these datasets publicly available, future
analysis can contribute to making Canada's transportation system more
just and fair, considering accessibility's as the main social good of
transportation \citep{martensTransportJusticeDesigning2016}, and the
inherent connection between public transit and the ``right to the city''
\citep{cogginRightTransportMoving2015}.
